\textit{Monami} as an oscillatory hydrodynamic instability in a submerged sea grass bed

### 2014-10-02 Paper/grass.tex
Cite White and Nepf, 2007

\maketitle
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production  \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
The simplest response of these flexible vegetation canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.
\newline
Current explanations of \monami ~invoke the existence of a shear layer due to vegetation drag  \cite{Ghisal02,Raupach96} and its instability through a mechanism similar to the Kelvin-Helmholtz instability. 
This instability is thought to lead to coherent eddies over the sea grass meadow, which then drive large amplitude synchronous oscillations.
\newline
While the shear layer model appears to predict the frequency of \monami ~consistent with experimental observations, several aspects of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis \cite{Raupach96} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds numbers \cite{drazin}, whereas observations in the lab \cite{Ghisal02} and the field \cite{Grizzle96} indicate the existence of a threshold flow speed below which \monami ~is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the top of the meadow, is in many cases comparable to the unvegetated layer thickness, and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model for the hydrodynamic linear stability analysis that accounts correctly for these effects, while also explaining lab experiments and field observations.
\newline
The drag exerted by the grass bed on the flow is central to the hypothesized instability for \monami. 
The instability and the resulting flow structures, though manifest in the waving of grasses, also persist if the grass blades are replaced by rigid dowels \cite{Ghisal02}. 
Therefore, for simplicity, we consider the grass bed to exchange momentum with the surrounding fluid but not deform. 
A criteria for the validity of this assumptions is presented at the end of this letter.
% Indeed, a more detailed calculation modeling terrestrial grass demonstrated \cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, the flow instability is similar to the instability of a shear layer.

@@

@@

This dependence of the boundary layer thickness on the vegetation density $N_g$, shown in fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.

@@

The growth rate $\sigma$ for a given wave number $k=2\pi /\lambda$ satisfies the eigenvalue problem.

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of $O(100)$, above which the flow is unstable (Re$(\sigma)>0$), emerges from the 
solution of eq.~\eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the vegetation density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model \cite{Raupach96}, or because of ad-hoc
assumptions for the mean velocity profile \cite{Raupach96,Delangre06}.

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified vegetation number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

>>>>

\maketitle
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production  \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. 
The simplest response of these flexible vegetation canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami ~\cite{AckermanOkubo93}.  
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.
Current explanations of \monami ~invoke the existence of a shear layer due to vegetation drag  \cite{Ghisal02,Raupach96} and its instability through a mechanism similar to the Kelvin-Helmholtz instability. 
This instability is thought to lead to coherent eddies over the sea grass meadow, which then drive large amplitude synchronous oscillations.
While the shear layer model appears to predict the frequency of \monami ~consistent with experimental observations, several aspects of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis \cite{Raupach96,White07} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds numbers \cite{drazin}, whereas observations in the lab \cite{Ghisal02} and the field \cite{Grizzle96} indicate the existence of a threshold flow speed below which \monami ~is not observed. 
(iii) The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the top of the meadow, is in many cases comparable to the unvegetated layer thickness, and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a mathematical model for the hydrodynamic linear stability analysis that accounts correctly for these effects, while also explaining lab experiments and field observations.
The drag exerted by the grass bed on the flow is central to the hypothesized instability for \monami. 
The instability and the resulting flow structures, though manifest in the waving of grasses, also persist if the grass blades are replaced by rigid dowels \cite{Ghisal02}. 
Therefore, for simplicity, we consider the grass bed to exchange momentum with the surrounding fluid but not deform. 
A criteria for the validity of this assumptions is presented at the end of this letter.
% Indeed, a more detailed calculation modeling terrestrial grass demonstrated \cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, the flow instability is similar to the instability of a shear layer. 

@@

\begin{figure*}
\begin{tabular}{cccc}
{\includegraphics[scale = 0.67]{Set4_dens28_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens30_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens32_imgsc}} &
{\includegraphics[scale = 0.67]{Set4_dens34_imgsc}} \\
{\includegraphics[scale = 0.67]{Set5_dens38_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens40_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens42_imgsc}} &
{\includegraphics[scale = 0.67]{Set5_dens46_imgsc}} \\
%{\includegraphics[scale = 0.67]{Set6_dens32_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens34_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens36_imgsc}} &
%{\includegraphics[scale = 0.67]{Set6_dens38_imgsc}}
\end{tabular}
\caption{$Re(\sigma)$ as function of wavenumber and Reynolds number for a specified vegetation number density $\Ndg$ along with the neutral curve ($Re(\sigma)$=0), for parameters shown in the corresponding panel.  
As $\Ndg \propto N_g$ increases, the unstable region splits into two labeled as ``Mode 1'' and ``Mode 2''. 
For $\Ndg$ below a critical value, the Mode 1 sets the threshold $\Rey$ whereas above the critical $\Ndg$ onset is determined by Mode 2.}
\label{K_Re_sigma_set3}
\end{figure*}

@@

The dependence of the boundary layer thickness on the vegetation density $N_g$, shown in fig.~\ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.

@@

The growth rate $\sigma$ for a given wave number $k=2\pi /\lambda$ satisfies the eigenvalue problem

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of $O(100)$, above which the flow is unstable (Re$(\sigma)>0$), emerges from the 
solution of eq.~\eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in fig.~\ref{Re_vs_delta}, and is found to compare well with experimental observations \cite{Ghisal02}.
The threshold Reynolds number increases with the vegetation density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
A similar conclusion was presented \cite{White07} in the case flow around emergent (i.e., $\hg>2H$) sea grass patch, but by assuming $U(y)$ to be a tanh-profile \textit{ad hoc}, and neglecting the viscous term.
Linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model \cite{Raupach96}, or because of the \textit{ad hoc}
assumptions about the mean velocity profile \cite{Raupach96,Delangre06}.
